### 2016-03-14 paper/icfp16-submission/gallery2.tex
misc edits everywhere

  let _ = wwhile (__f__, 2)
\end{ecode}
%
Unfortunately, the student has forgotten to \emph{apply}
@f@ at all on line 2, and just matches it directly against
a pair.
This faulty definition of @wwhile@ still typechecks however,
and hence, are ``assumed'' to be correct by and both \ocaml\
and \sherrloc\ which blame the \emph{use-site} on line 10.
%
\begin{verbatim}
  This expression has type
    int -> int * bool
  but an expression was expected of type
    'a * bool
\end{verbatim}
%
\toolname\ synthesizes a trace that draws the eye to the
true error: the @match@ expression on line 2, and highlights
the conflict in matching a function against a pair pattern.
%
\begin{center}
  \includegraphics[height=150px]{wwhile.png}
\end{center}
%
By highlighting conflicting values (\ie\ the source and sink
of the problem) and not making any assumptions, \toolname\
focusses the user's attention on the piece of code that is
actually relevant to the error.

>>>>

  let _ = wwhile (__f__, 2)
\end{ecode}
%
Unfortunately, the student has forgotten to \emph{apply}
@f@ at all on line 2, and just matches it directly against
a pair.
This faulty definition of @wwhile@ still typechecks however,
and is assumed to be correct by and both \ocaml\
and \sherrloc\ which blame the use-site on line 10.
%
\begin{verbatim}
  This expression has type
    int -> int * bool
  but an expression was expected of type
    'a * bool
\end{verbatim}
%
\toolname\ synthesizes a trace that draws the eye to the
true error: the @match@ expression on line 2, and highlights
the conflict in matching a function against a pair pattern.
%
\begin{center}
  \includegraphics[height=150px]{wwhile.png}
\end{center}
%
By highlighting conflicting values (\ie\ the source and sink
of the problem) and not making any assumptions, \toolname\
focusses the user's attention on the piece of code that is
actually relevant to the error.

### 2016-03-14 paper/icfp16-submission/overview.tex
misc edits everywhere

\paragraph{Jump-compressed Witnesses}
It is rare for the initial state of the visualization to be
informative enough to diagnose the error.
%
We provide a short-cut to expand the witness to a \emph{jump-compressed}
trace, which contains every function call and return step.
%
The jump-compressed trace abstracts the computation as a sequence of
call-response pairs, providing a high-level overview of steps taken
to reach the crash, and a high level of compression compared to the
full trace.
%
For example, the jump-compressed trace in Figure~\ref{fig:factorial}
contains 4 nodes compared to the 19 in the fully expanded trace.
%
Our benchmark suite of student programs shows that jump-compression is
practical, with an average trace size of 7 nodes and a median of 5.

>>>>

\paragraph{Jump-compressed Witnesses}
It is rare for the initial state of the visualization to be
informative enough to diagnose the error.
%
Rather than abandon the user, we provide a short-cut to expand the witness
to a \emph{jump-compressed} trace, which contains every function call
and return step.
%
The jump-compressed trace abstracts the computation as a sequence of
call-response pairs, providing a high-level overview of steps taken
to reach the crash, and a high level of compression compared to the
full trace.
%
For example, the jump-compressed trace in Figure~\ref{fig:factorial}
contains 4 nodes compared to the 19 in the fully expanded trace.
%
Our benchmark suite of student programs shows that jump-compression is
practical, with an average jump-compressed trace size of 7 nodes and a
median of 5.

### 2016-03-14 paper/icfp16-submission/semantics.tex
misc edits everywhere

\subsection{Semantics}
\label{sec:semantics}
%
Recall that our goal is to synthesize a value that demonstrates
why (and how) a function goes wrong.
%
We accomplish this by combining evaluation with type inference,
giving us a form of dynamic type inference.
% starting with an unconstrained hole and gradually refining it
% as evaluation constrains its type.
%
Each primitive evaluation step tells us more about the types of the
program values. For example, addition tells us that the addends must be
integers, and case-analysis tells us that the scrutinee must be a tree.
%
When a hole appears in such a context, we know what type it must have
in order to make progress and can fill it in with a concrete value.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen}, takes
as input a type $t$ and returns a value of that type.
%
For base types, the procedure returns an arbitrary value of
that type.
%
For functions, it returns a lambda with a \emph{new} hole
denoting the return value, and unconstrained types (denoted
by $\thole$) yield fresh holes constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
When generating a $\ttree{t}$ we must take care to ensure
the resulting tree is well-typed.
%
For a polymorphic type $\ttree{\thole}$ %, eg in \recasegoodone,
we will place holes in the generated nodes; they will be lazily filled
in later, on demand.

@@

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type, under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a lambda-term $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} concrete type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any lambda-term $f$, if
  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists a $v$ of type $t$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$.
\end{thm}

We need to develop some machinery in order to prove this theorem.
First, we show how our evaluation rules encode a dynamic form of
type inference, and next, we show that the types inferred via
evaluation are indeed maximally general.

\paragraph{The Type of a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck, \eg
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) data types are obtained from the
labels on their values.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}}
as synthesizing a partial instantiation of \thole, and hence,
\emph{dynamically inferring} a (partial) type for $f$'s input.
%
We can extract this type from an evaluation trace by
applying the final type substitution to \thole.
% \emph{resolving} the \thole\ with the final type
% substitution at the end of the trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ up to $\trace$, written
\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.

@@

We prove that the partial instantiation of a type hole $\thole$
is always a refinement of the partial instantiation of the
associated value hole $\vhole{\thole}$.

@@

\paragraph{Incompatible Types Are Wrong}
\emph{For all} types that are \emph{incompatible} with the
partial input type up to trace $\trace$, there exists a value
that will cause $f$ to get stuck in \emph{at most} $k$ steps,
where $k$ is the length of $\trace$.

@@

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
So far, we have seen how a trace leading to a stuck configuration yields
a general witness demonstrating that the program is ill-typed (\ie\ goes
wrong for at least one input of every type.)
In particular, we have shown how to non-deterministically find a witnesses
for a function of a \emph{single} argument.

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $k$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, the target expression $e$ that contains the type error,
and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $n$ traces.

@@

\paragraph{Generating Witnesses}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure
$\genWitness{k}{e}$, which takes as input a bound $k$ and the
target expression $e$, and returns a list of witness expressions
$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
gets stuck.
%
The search proceeds as follows.
%
\begin{enumerate}
  \item We invoke $\genArgs{e}$ to produce a \emph{saturated}
        application $e_{sat}$.

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu}\ldots$$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ to be
  the uncurried version of $f$ that takes all of its arguments as a
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}

>>>>

\subsection{Semantics}
\label{sec:semantics}
%
Recall that our goal is to synthesize a value that demonstrates
why (and how) a function goes wrong.
%
We accomplish this by combining evaluation with type inference,
giving us a form of dynamic type inference.
% starting with an unconstrained hole and gradually refining it
% as evaluation constrains its type.
%
Each primitive evaluation step tells us more about the types of the
program values. For example, addition tells us that the addends must be
integers, and % case-analysis tells us that the scrutinee must be a tree.
an if-expression tells us the condition must be a boolean.
%
When a hole appears in such a context, we know what type it must have
in order to make progress and can fill it in with a concrete value.

@@

\paragraph{Generating Values} The (non-deterministic)
$\gen{t}{\tsu}$ in Figure~\ref{fig:gen} takes
as input a type $t$ and returns a value of that type.
%
For base types the procedure returns an arbitrary value of
that type.
%
For functions it returns a lambda with a \emph{new} hole
denoting the return value.
%
For unconstrained types (denoted
by $\thole$) it yields a fresh hole constrained to have type
\thole (denoted by $\vhole{\thole}$).
%
When generating a $\ttree{t}$ we must take care to ensure
the resulting tree is well-typed.
%
For a polymorphic type $\ttree{\thole}$ %, eg in \recasegoodone,
or $\tprod{\thole_1}{\thole_2}$
we will place holes in the generated value; they will be lazily filled
in later, on demand.

@@

A key technical challenge in generating witnesses is
that we have no (static) type information to rely upon.
%
Thus, we must avoid the trap of generating \emph{spurious}
witnesses that arise from picking irrelevant values, when
instead there exist perfectly good values of a \emph{different}
type under which the program would not have gone wrong.
%
We now show that our evaluation relation instantiates holes
in a \emph{general} manner. That is, given a lambda-term $f$,
if we have $\steps{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,
then \emph{for every} concrete type $t$, we can find a value
$v$ of type $t$ such that $\eapp{f}{v}$ goes wrong.

\begin{thm}{\textbf{[Witness Generality]}}
\label{thm:soundness}
  For any lambda-term $f$, if
  \hbox{$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\stuck}{\vsu}{\tsu}$,}
  then for every % inhabitable
  type
  % \footnote{We exclude builtin functions that subvert the type system, \eg \texttt{Obj.magic}, and thus consider the type $\forall a b. \tfun{a}{b}$ to be uninhabitable.}
  $t$ there exists a value $v$ of type $t$ such that
  $\steps{\eapp{f}{v}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{thm}

We need to develop some machinery in order to prove this theorem.
First, we show how our evaluation rules encode a dynamic form of
type inference, and then we show that the witnesses found by
evaluation are indeed maximally general.

\paragraph{The Type of a Value} The \emph{dynamic type}
of a value $v$ is defined as a function $\typeof{v}$ shown
in Figure~\ref{fig:typeof}.
%
The types of primitive values are defined in the natural manner.
%
The types of functions are \emph{approximated}, which is all
that is needed to ensure an application does not get stuck.
%
For example,
$$\typeof{\efun{x}{\eplus{x}{1}}} = \tfun$$
instead of $\tint \rightarrow \tint$.
%
The types of (polymorphic) trees are obtained from the labels on their
values, and the types of tuples directly from their values.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\vhole{\thole}}
as synthesizing a partial instantiation of \thole, and thus
\emph{dynamically inferring} a (partial) type for $f$'s input.
%
We can extract this type from an evaluation trace by
applying the final type substitution to \thole.
% \emph{resolving} the \thole\ with the final type
% substitution at the end of the trace.
%
Formally, we say that if
$\steptr{\trace}{\eapp{f}{\vhole{\thole}}}{\emptysu}{\emptysu}{\estuck}{\vsu}{\tsu}$,
then the \emph{partial input type} of $f$ up to $\trace$, written
\ptype{\trace}{f}, is $\resolve{\thole}{\tsu}$.

@@

% We prove that the partial instantiation of a type hole $\thole$
% is always a refinement of the partial instantiation of the
% associated value hole $\vhole{\thole}$.

@@

\paragraph{Incompatible Types Are Wrong}
\emph{For all} types that are \emph{incompatible} with the
partial input type up to $\trace$, there exists a value
that will cause $f$ to get stuck in \emph{at most} $k$ steps,
where $k$ is the length of $\trace$.

@@

\subsection{Search Algorithm}
\label{sec:search-algorithm}
%
So far, we have seen how a trace leading to a stuck configuration yields
a general witness demonstrating that the program is ill-typed (\ie\ goes
wrong for at least one input of every type).
In particular, we have shown how to non-deterministically find a witnesses
for a function of a \emph{single} argument.

@@

\paragraph{Inputs and Outputs}
%
The problem of generating inputs is undecidable in general.
%
Our witness generation procedure takes two inputs.
%
First, a search bound $k$ which is used to define the \emph{number} of
traces to explore. (We assume, without loss of generality, that all
traces are finite.)
%
Second, the target expression $e$ that contains the type error,
and may be a curried function of multiple arguments.
%
The witness generation procedure returns as output a list of (general)
witness expressions, each of which is of the form $e\ v_1 \ldots v_n$.
%
The \emph{empty} list is returned when no witness can be found after
exploring $k$ traces.

@@

\paragraph{Generating Witnesses}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses with the procedure
$\genWitness{k}{e}$, which takes as input a bound $k$ and the
target expression $e$, and returns a list of witness expressions
$\eapp{e}{v_1 \ldots v_n}$ that demonstrate how the input program
gets stuck.
%
The search proceeds as follows.
%
\begin{enumerate}
  \item We invoke $\genArgs{e}$ to produce a \emph{saturated}
        application $e_{sat}$.

@@

\begin{cor}{\textbf{[Witness Generation]}}
\label{thm:generation}
  If $$\genWitness{k}{e} = \triple{\eapp{e}{v_1 \ldots v_n}}{\vsu}{\tsu},\ldots$$
  then for all types $t_1 \ldots t_n$ there exist values $w_1 \ldots w_n$ such that
  $\steps{\eapp{e}{w_1 \ldots w_n}}{\emptysu}{\emptysu}{\stuck}{\vsu'}{\tsu'}$.
\end{cor}
\begin{proof}
  For any function $f$ of multiple arguments, we can define $f'$ to be
  the uncurried version of $f$ that takes all of its arguments as a single
  nested pair, and then apply Theorem~\ref{thm:soundness} to $f'$.
\end{proof}
